The dynamics of starvation and recovery

### 2016-08-05 Manuscript/7.tex
Yeakel version of ms, and updated figures for ef

%\usepackage{scicite}
%\usepackage{times}
%\usepackage{graphicx}
%\usepackage{verbatim}
%\usepackage{subfigure}
%\usepackage{tikz}
%\usepackage{amsmath}
%\usepackage{amssymb}
%\usepackage{array}
%\usepackage{tabularx}
%\usepackage{multirow}
%\usepackage{multicol}
%\usepackage{multibox}
%\usepackage{rotating}
%\usepackage{layout}
%\usepackage{epstopdf}
%\usepackage{afterpage}
%\usepackage[left]{lineno}

@@

%\usepackage{amssymb,amsfonts,amsmath}
\usepackage{amsmath}

@@

\begin{abstract}
This is the abstract.  \emph{ \sid{No it isn't.  It's merely a placeholder.}}
\end{abstract}
\keywords{foraging | starvation|reproduction}

@@

%Minimal mass bound
A lower bound on mammalian body size is given by $\epsilon=1$, where mammals
have no metabolic reserves and immediately starve; this occurs at a size of
\sid{$M=value$}.  This calculation \sid{what calculation?} gives an extreme
limit on size but does not account for the subtleties of starvation dynamics
that may limit body size. The NSM correctly predicts that species with
smaller masses have larger steady-state population densities.  However we
observe that there is a sharp change in the mass dependence of both the
steady-state densities and $\sigma/\rho$ at $M \approx 0.3$ grams
(Fig.~\ref{fig:mass}a,b).  The dependence of the rates of starvation and
recovery explain this phenomenon.  As the mass decreases, the rate of
starvation increases, while the rate of recovery declines super-exponentially
\sid{how do we know this?}.  This decline in $\rho$ occurs when the
percentage of body fat is
$1 - 1/\left[\big(a/b\big)^{1/\left(\eta-1\right)}M\right] \approx
2\%$,
whereupon consumers have no eligible route \sid{what does this mean?} that avoids
starvation.  Compellingly, this dynamic bound determined by the rate of
energetic recovery is close to the minimum observed mammalian body size of
ca. 1.3-2.5 grams (Fig.~\ref{fig:mass}b,c), a range that occurs as the
recovery rate begins its decline.  In addition to known transport
limitations~\cite{West:2002ud}, we suggest that an additional constraint of
lower body size stems from the dynamics of starvation.  This work \sid{which
  work?} mirrors other efforts where coincident limitations seem to limit the
smallest possibilities for life within a particular class or
organisms~\cite{Kempes:2012hy}.

@@

@@

\bibliography{aa_starving}

\begin{acknowledgments}
  J.D.Y. was supported by an Omidyar postdoctoral Fellowship at the Santa Fe Institute. 
  C.P.K was supported by a Trump Fellowship from the American League of
  Conservatives.  
  S.R. was supported by grants DMR-1608211 and 1623243 from
  the National Science Foundation, and by the John Templeton Foundation.
\end{acknowledgments}

>>>>

%\usepackage{scicite}
%\usepackage{times}
%\usepackage{graphicx}
%\usepackage{verbatim}
%\usepackage{subfigure}
%\usepackage{tikz}
%\usepackage{array}
%\usepackage{tabularx}
%\usepackage{multirow}
%\usepackage{multicol}
%\usepackage{multibox}
%\usepackage{rotating}
%\usepackage{layout}
%\usepackage{epstopdf}
%\usepackage{afterpage}
%\usepackage[left]{lineno}

@@

%\usepackage{amssymb,amsfonts,amsmath}
\usepackage{amsmath}
\usepackage{amssymb}

@@

\begin{abstract}
We introduce a minimal foraging model that incorporates two classes of
foragers: nutritionally replete foragers that can procreate, and
undernourished and non-procreating foragers that are susceptible to
mortality.  As a function of the transition rates between these replete and
undernourished states that are determined by the presence or absence of
resources, as well as the replenishment rate of the underlying resource, the
forager populations can either undergo cyclic dynamics or reach a steady
state.  By appplying basic allometric considerations, we obtain strong
constraints on these transition rates and thereby find that the population
dynamics that is subject to these constraints is typically driven to a steady
state.  As a further consequence of this viewpoint, we find that
allometrically constrained rates fall within a `refuge' in parameter space,
where the probability of extinction of the consumer population is minimized.
Thus our model provides a mechanism that may both drive and constrain natural
animal populations.  We also find that our model provides a natural framework
to determine a natural size for animal populations by examining the relative
stability of a otherwise homoegeneous population to a mutant population of a
different size.  For body masses $\lesssim 10^6$g, a more massive invader
dominates over the resident population, and vice versa for body mass $\gtrsim
10^6$g, thus providing a simple principled mechanism for Cope's rule.

\end{abstract}
\keywords{foraging | starvation|reproduction}

@@

%Minimal mass bound
A lower bound on mammalian body size is given by $\epsilon=1$, where mammals
have no metabolic reserves and immediately starve; this occurs at a size of
\sid{$M=value$}.  This calculation \sid{what calculation?} gives an extreme
limit on size but does not account for the subtleties of starvation dynamics
that may limit body size. The NSM correctly predicts that species with
smaller masses have larger steady-state population densities.  However we
observe that there is a sharp change in the mass dependence of both the
steady-state densities and $\sigma/\rho$ at $M \approx 0.3$ grams
(Fig.~\ref{fig:mass}a,b).  The dependence of the rates of starvation and
recovery explain this phenomenon.  As the mass decreases, the rate of
starvation increases, while the rate of recovery declines super-exponentially
\sid{how do we know this?}.  This decline in $\rho$ occurs when the
percentage of body fat is
$1 - \big(a/b\big)^{-1/\left(\eta-1\right)}M^{-1} \approx
2\%$,
whereupon consumers have no eligible route \sid{what does this mean?} that avoids
starvation.  Compellingly, this dynamic bound determined by the rate of
energetic recovery is close to the minimum observed mammalian body size of
ca. 1.3-2.5 grams (Fig.~\ref{fig:mass}b,c), a range that occurs as the
recovery rate begins its decline.  In addition to known transport
limitations~\cite{West:2002ud}, we suggest that an additional constraint of
lower body size stems from the dynamics of starvation.  This work \sid{which
  work?} mirrors other efforts where coincident limitations seem to limit the
smallest possibilities for life within a particular class or
organisms~\cite{Kempes:2012hy}.

@@

A potential critique of our results, which show that larger mammals are less susceptible to extinction, is why we don't observe a greater number of large mammals in the modern world. However, recent research suggests that the pleistocene may have been much more populated with a significant diversity of very large mammals \cite{Doughty:2013kd,Doughty:2015hy,Doughty:2015je} which where also much more geographically widespread than today. These results combined with our findings suggest that the modern diversity of mammals may not represent a true steady state the current distribution of nutrients and large seeds may be very different from the past \cite{Doughty:2013kd,Doughty:2015hy,Doughty:2015je}.

@@

\bibliography{aa_starving2}

\begin{acknowledgments}
  J.D.Y. was supported by an Omidyar postdoctoral Fellowship at the Santa Fe
  Institute and by startup funds at the University of California, Merced.
  C.P.K was supported by an Omidyar postdoctoral Fellowship at the Santa Fe
  Institute.  S.R. was supported by grants DMR-1608211 and 1623243 from the
  National Science Foundation, and by the John Templeton Foundation, all at
  the Santa Fe Institute.
\end{acknowledgments}

### 2016-08-05 Manuscript/8.tex
Yeakel version of ms, and updated figures for ef

\begin{abstract}
We introduce a minimal foraging model that incorporates two classes of
foragers: nutritionally replete foragers that can procreate, and
undernourished and non-procreating foragers that are susceptible to
mortality.  As a function of the transition rates between these replete and
undernourished states that are determined by the presence or absence of
resources, as well as the replenishment rate of the underlying resource, the
forager populations can either undergo cyclic dynamics or reach a steady
state.  By appplying basic allometric considerations, we obtain strong
constraints on these transition rates and thereby find that the population
dynamics that is subject to these constraints is typically driven to a steady
state.  As a further consequence of this viewpoint, we find that
allometrically constrained rates fall within a `refuge' in parameter space,
where the probability of extinction of the consumer population is minimized.
Thus our model provides a mechanism that may both drive and constrain natural
animal populations.  We also find that our model provides a natural framework
to determine a natural size for animal populations by examining the relative
stability of a otherwise homoegeneous population to a mutant population of a
different size.  For body masses $\lesssim 10^6$g, a more massive invader
dominates over the resident population, and vice versa for body mass $\gtrsim
10^6$g, thus providing a simple principled mechanism for Cope's rule.

\end{abstract}
\keywords{foraging | starvation|reproduction}

@@

From Eq.~\eqref{eq:ss}, an obvious constraint on the NSM is that the
reproduction rate $\lambda$ must be less than the starvation rate $\sigma$,
so that $R^*$ is positive.  In fact, when the resource density $R=0$, the
rate equation for $F$ gives exponential growth of $F$ for $\lambda>\sigma$.
The condition $\sigma = \lambda$ represents a transcritical (TC) bifurcation
that demarcates the physical and unphysical regimes\sid{give Ref}.  The
biological implication of the constraint $\lambda<\sigma$ has a simple
interpretation---the rate at which a macroscopic organism loses mass due to
lack of resources is generally much faster than the rate of reproduction.  As
we will discuss below, this inequality is a natural consequence of allometric
constraints~\cite{Kempes:2012hy} for organisms within empirically observed
body size ranges (Fig.~\ref{fig:gvs}).

@@

Whereas the relation between consumer growth rate $\lambda$ and the
starvation rate $\sigma$ defines an absolute bound of biological
feasibility---the TC bifurcation---the starvation rate $\sigma$ also
determines the sensitivity of the consumer population to changes in resource
density.  When $\sigma\gg\lambda$, the steady-state population density is
small, thereby increasing the risk of stochastic extinction.  On the other
hand, as $\sigma$ decreases, the system will ultimately be poised either near
the TC or the Hopf bifurcation (Fig.~\ref{fig:fp}).  If the recovery rate
$\rho$ is sufficiently small, the TC bifurcation is reached and the resource
eventually is eliminated.  If $\rho$ exceeds a threshold value, cyclic
dynamics will develop as the Hopf bifurcation is approached.

\sid{mention  and refer to the starving random walk model somewhere.}
\\

@@

Previous studies have used allometric constraints to explain the periodicity
of cyclic populations
\cite{CalderIII:1983jd,Peterson:1984hj,Krukonis:1991fk}, suggesting a period
$\propto M^{0.25}$, however this relation seems to hold only for some species
\cite{Hendriks:2012fc} and competing explanations [related to ....\sid{???}]
exist~\cite{Kendall:1999iy,Hogstedt:2005cr}.
%It is more difficult to ascertain whether a population is truly cycling according to an internal or external dynamic~\cite{Kendall:1999iy} or if fluctuations are due to stochasticity, and this is particularly true for longer-lived organisms.
Statistically significant support for the existence of population cycles
among mammals is predominantly based on time series for small
mammals~\cite{Kendall:1998hl}, where we our model would predict much longer
and more pronounced transient dynamics, given how close these points are to
the Hopf bifurcation.  On the other hand, the longer gestational times and
the increased difficulty in measurements, precludes obtaining similar-quality
data for larger organisms.
\\
%, we acknowledge that absence of evidence for cyclic dynamics among longer-lived species is not necessarily evidence of absence.
 %, emphasizing the role of starvation on dynamics.

%Extinction to the left and the right

\noindent {\bf Extinction risk}\\
Within our model, higher rates of starvation result in a larger flux of the
population to the hungry state.  In this state reproduction is absent, thus
increasing the likelihood of extinction.  However, from the perspective of
population survival, it is the rate of starvation relative to the rate of
recovery that determines the long-term dynamics of the system
(Fig.~\ref{fig:fp}).  We now examine the competing effects of cyclic dynamics
vs.\ changes in steady state density on extinction risk as a function of the
ratio $\sigma/\rho$.  To this end, we computed the probability of extinction,
where extinction is defined as the population trajectory going below
$0.2~\times$ the allometrically constrained steady state for all times
between $10^2$ and $\leq 10^6$.  This procedure is repeated for 1000
replicates of the continuous-time system shown in Eq.~\ref{eq:system} for an
organism of $M=100$ grams.  In each replicate the initial condition is
distributed around the steady state (Eq.~\ref{eq:ss}).  Specifically the
initial densities are chosen to be $A(F^*,H^*,R^*)$, with $A$ a random
variable that is uniformly distribution in $[0,2]$.  By allowing the rate of
starvation to vary, we assessed extinction risk across a range of values of
the ratio $\sigma/\rho$ varying between $10^{-2}$ to $2.5$, thus examining a
horizontal cross-section of Fig.~\ref{fig:fp}.  As expected, higher rates of
extinction correlated with both low and high values of $\sigma/\rho$.  For
low values of $\sigma/\rho$, the increased extinction risk results from
transient cycles with larger amplitudes as the system nears the Hopf
bifurcation (Fig.~\ref{fig:ext}).  For large values of $\sigma/\rho$, higher
extinction risk arises because of to the decrease in the steady state
consumer population density.  This interplay creates an `extinction refuge'
as shown in Fig.~\ref{fig:ext}, such that for a relatively constrained range
of $\sigma/\rho$, extinction probabilities are minimized.

@@

\bibliography{aa_starving2}

\begin{acknowledgments}
  J.D.Y. was supported by an Omidyar postdoctoral Fellowship at the Santa Fe
  Institute and by startup funds at the University of California, Merced.
  C.P.K was supported by an Omidyar postdoctoral Fellowship at the Santa Fe
  Institute.  S.R. was supported by grants DMR-1608211 and 1623243 from the
  National Science Foundation, and by the John Templeton Foundation, all at
  the Santa Fe Institute.
\end{acknowledgments}

>>>>

\begin{abstract} %246 words
The eco-evolutionary dynamics of species is fundamentally linked to the energetic constraints of individuals.
Of particular importance are the tradeoffs between reproduction and the dynamics of starvation and recovery in resource limited environments.
We introduce a minimal nutritional state-structured model that incorporates two classes of consumer: nutritionally replete consumers that reproduce, and
undernourished, non-reproducing consumers that are susceptible to
mortality.  
As a function of the energetic rates between these replete and
undernourished states that are determined by the presence or absence of
resources, the consumer populations can either undergo cyclic dynamics or reach a steady state.
We obtain strong constraints on starvation and recovery rates by deriving allometric scaling relationships and find that population dynamics subject to these constraints can approach the cyclic regime but are typically driven to a steady state.
Moreover, we find these rates fall within a `refuge' in parameter space, where the probability of extinction of the consumer population is minimized.
Thus we introduce a potential mechanism that may both drive and constrain the dynamics of animal populations.
Importantly, our model provides a natural framework that predicts maximum body size for mammals by examining the relative stability of an otherwise homogeneous population to a mutant population with altered percent body fat.
For body masses $\lesssim 10^6$g, an invader with increased percent body fat dominates over the resident population, and vice versa for body mass $\gtrsim 10^6$g, thus providing a principled mechanism for a within-lineage driver of Cope's rule.
\end{abstract}
\keywords{foraging | starvation|reproduction}

@@

From Eq.~\eqref{eq:ss}, an obvious constraint on the NSM is that the
reproduction rate $\lambda$ must be less than the starvation rate $\sigma$,
so that $R^*$ is positive.  In fact, when the resource density $R=0$, the
rate equation for $F$ gives exponential growth of $F$ for $\lambda>\sigma$.
The condition $\sigma = \lambda$ represents a transcritical (TC) bifurcation
that demarcates the physical and unphysical regimes.  The
biological implication of the constraint $\lambda<\sigma$ has a simple
interpretation---the rate at which a macroscopic organism loses mass due to
lack of resources is generally much faster than the rate of reproduction.  As
we will discuss below, this inequality is a natural consequence of allometric
constraints~\cite{Kempes:2012hy} for organisms within empirically observed
body size ranges (Fig.~\ref{fig:gvs}).

@@

Whereas the relation between consumer growth rate $\lambda$ and the
starvation rate $\sigma$ defines an absolute bound of biological
feasibility---the TC bifurcation---$\sigma$ also
determines the sensitivity of the consumer population to changes in resource
density.  When $\sigma\gg\lambda$, the steady-state population density is
small, thereby increasing the risk of stochastic extinction.  On the other
hand, as $\sigma$ decreases, the system will ultimately be poised either near
the TC or the Hopf bifurcation (Fig.~\ref{fig:fp}).  If the recovery rate
$\rho$ is sufficiently small, the TC bifurcation is reached and the resource
eventually is eliminated.  If $\rho$ exceeds a threshold value, cyclic
dynamics will develop as the Hopf bifurcation is approached.
\\

@@

Previous studies have used allometric constraints to explain the periodicity
of cyclic populations
\cite{CalderIII:1983jd,Peterson:1984hj,Krukonis:1991fk}, suggesting a period
$\propto M^{0.25}$, however this relation seems to hold only for some species
\cite{Hendriks:2012fc} and competing explanations [related to ....\sid{???}]
exist~\cite{Kendall:1999iy,Hogstedt:2005cr}.
%It is more difficult to ascertain whether a population is truly cycling according to an internal or external dynamic~\cite{Kendall:1999iy} or if fluctuations are due to stochasticity, and this is particularly true for longer-lived organisms.
Statistically significant support for the existence of population cycles
among mammals is predominantly based on time series for small
mammals~\cite{Kendall:1998hl}, where we our model would predict much longer
and more pronounced transient dynamics, given how close these points are to
the Hopf bifurcation.  On the other hand, the longer gestational times and
the increased difficulty in measurements, precludes obtaining similar-quality
data for larger organisms.
\vspace{6mm}
%, we acknowledge that absence of evidence for cyclic dynamics among longer-lived species is not necessarily evidence of absence.
 %, emphasizing the role of starvation on dynamics.

%Extinction to the left and the right
\noindent {\bf Extinction risk}\\
Within our model, higher rates of starvation result in a larger flux of the
population to the hungry state.  In this state reproduction is absent, thus
increasing the likelihood of extinction.  However, from the perspective of
population survival, it is the rate of starvation relative to the rate of
recovery that determines the long-term dynamics of the system
(Fig.~\ref{fig:fp}).  We now examine the competing effects of cyclic dynamics
vs.\ changes in steady state density on extinction risk as a function of the
ratio $\sigma/\rho$.  To this end, we computed the probability of extinction,
where extinction is defined as the population trajectory going below
$0.2~\times$ the allometrically constrained steady state for all times
between $10^2$ and $\leq 10^6$.  This procedure is repeated for 1000
replicates of the continuous-time system shown in Eq.~\ref{eq:system} for an
organism of $M=100$ grams.  
In each replicate the initial condition is
distributed around the steady state (Eq.~\ref{eq:ss}).  Specifically the
initial densities are chosen to be $A(F^*,H^*,R^*)$, with $A$ a random
variable that is uniformly distribution in $[0,2]$.  By allowing the rate of
starvation to vary, we assessed extinction risk across a range of values of
the ratio $\sigma/\rho$ varying between $10^{-2}$ to $2.5$, thus examining a
horizontal cross-section of Fig.~\ref{fig:fp}.  
As expected, higher rates of
extinction correlated with both low and high values of $\sigma/\rho$.  For
low values of $\sigma/\rho$, the increased extinction risk results from
transient cycles with larger amplitudes as the system nears the Hopf
bifurcation (Fig.~\ref{fig:ext}).  For large values of $\sigma/\rho$, higher
extinction risk arises because of to the decrease in the steady state
consumer population density.  This interplay creates an `extinction refuge'
as shown in Fig.~\ref{fig:ext}, such that for a relatively constrained range
of $\sigma/\rho$, extinction probabilities are minimized.

@@

\bibliography{aa_starving3}

\begin{acknowledgments}
We thank Luis Bettancourt, Eric Libby, Seth Newsome, et al. for helpful discussions and comments that greatly increased the quality of the manuscript.
  J.D.Y. was supported by startup funds at the University of California, Merced, and an Omidyar postdoctoral Fellowship at the Santa Fe Institute.
  C.P.K. was supported by an Omidyar postdoctoral Fellowship at the Santa Fe
  Institute.  
  S.R. was supported by grants DMR-1608211 and 1623243 from the National Science Foundation, and by the John Templeton Foundation, all at the Santa Fe Institute.
\end{acknowledgments}
